pipeComp, a general framework for the evaluation of computational pipelines, reveals performant single-cell RNA-seq preprocessing tools

### 2020-03-01 draft.tex
Updates from Overleaf

Here, we develop a flexible R framework for pipeline comparison and evaluate the various steps of analysis leading from an initial count matrix to a cluster assignment, which are critical in a wide range of applications. We collected real datasets of known cell composition (Table \ref{tab:table1}) and used a variety of evaluation metrics to investigate in a multilevel fashion the impact of various parameters and variations around a core scRNAseq pipeline. Although we use some datasets based on other protocols, our focus is especially on droplet-based datasets that do not include exogenous control RNA (i.e. spike-ins); see Table \ref{tab:table1} and Figure \ref{fig:figure1} for more details. In addition to previously-used benchmark datasets with true cell labels \cite{duoClustering2018,tianMixology2018}, we simulated two datasets with a hierarchical subpopulation structure based on real 10x human and mouse data using \textit{muscat} \cite{CrowellMuscat2019}. 
Since graph-based clustering \cite{satijaSeurat2015} was previously shown to consistently perform well across several datasets \cite{duoClustering2018,tianMixology2018}, we used the \textit{Seurat} pipeline as the starting point to perform an integrated investigation of: 1) doublet identification, 2) cell filtering, 3) normalization, 4) feature selection, 5) dimension reduction, 6) clustering. We compared competing approaches and also explored more fine-grained parameters and variations on common methods. Importantly, the success of methods at a certain analytical step might be dependent on choices at other steps. Therefore, instead of evaluating each step in isolation, we developed a general framework for evaluating nested variations on a pipeline and suggest a multilevel panel of metrics. Finally, we evaluate several recent methods and provide concrete recommendations.

@@

Doublets, defined as two cells sequenced under the same cellular barcode (e.g., being captured in the same droplet), are fairly frequent in scRNAseq datasets, with estimates ranging from 1 to 10\% depending on the platform and cell concentration used \cite{bloomEstimating2018,kangMultiplexedDemuxlet2018}. While doublets of the same cell type are relatively innocuous for most downstream applications due to their conservation of the relative expression between genes, doublets formed from different cell types or states are likely to be misclassified and could potentially distort downstream analysis. In some cases, doublets can be identified through their unusually high number of reads and detected features, but this is not always the case (Supplementary Figure 2). A number of methods were developed to identify doublets, in most cases by comparing each cell to artificially-created doublets \cite{mcginnisDoubletfinder2019, LunScran2016, BaisScds2019}. We first evaluated the capacity of these methods to detect doublets using the two 10x datasets with cells of different genetic identity \cite{tianMixology2018}, using SNP genotypes as the ground truth. For the sole purpose of this section, we included an additional dataset with SNP information but lacking true cell labels \cite{kangMultiplexedDemuxlet2018}. Of note, SNP-based analyses also call doublets created by cells of the same cell type (but from different individuals) and which are generally described as homotypic (as opposed to neotypic or heterotypic doublets, i.e. doublets from different cell types). These homotypic doublets might not be identifiable from the mere gene counts, and their identification is not generally the primary aim of doublet callers since they are often considered innocuous and, when across individuals, can be identified through other means (e.g., SNPs). We therefore do not expect a perfect accuracy in datasets involving cells of the same type across individuals (as in the demuxlet dataset).

@@

We next investigated the impact of filtering according to various criteria (see Methods). An important risk of excluding cells on the basis of their distance from the whole distribution of cells on some properties (e.g., library size) is that these properties tend to have different distributions across subpopulations. As a result, thresholds in terms of number of MADs from the whole distribution can lead to strong biases against certain subpopulations (Figure \ref{fig:figure4}A). We therefore examined the tradeoff between the increased accuracy of filtering and the maximum proportion of cells excluded per subpopulation (Figure \ref{fig:figure4}B and Supplementary Figure 8). Since filtering changes the relative abundance of the different subpopulations, global clustering accuracy metrics such as ARI are not appropriate here. We therefore calculated the per-subpopulation precision and recall using the Hungarian algorithm \cite{PapadimitriouHu1998} and monitored their mean F1 score. A first observation was that, although the stringent filtering tended to be associated with an increase in accuracy, it could also become deleterious and most of the benefits could be achieved without very stringent filtering and minimizing subpopulation bias (Figure \ref{fig:figure4}B). Applying the same filtering criteria on individual clusters of cells (identified through scran's \texttt{quickCluster} method) resulted in nearly no cell being filtered out. This suggests that filtering on the global population tends to discard cells of subpopulations with more extreme properties (e.g. high library size), rather than low-quality cells. Finally, by changing the distributions, the doublet removal step in conjunction with filtering sometimes resulted in a net decrease in the proportion of excluded cells while retaining or improving accuracy. We therefore recommend the use of doublet removal followed by relatively mild filtering, such as is implemented in our `default' filtering (see Methods).

@@

We compared the various estimates of dimensionality in their ability to retrieve the intrinsic number of dimensions in a dataset, based on Seurat's weighted PCA space. As a first approximation of the true dimensionality, we computed the variance in each principal component that was explained by the subpopulations, which sharply decreased after the first few components in most datasets (Figure \ref{fig:figure8}A). Figure \ref{fig:figure8}B shows the difference between the dimension estimates of the above methods and that based on the subpopulations (i.e. from Figure \ref{fig:figure8}A). Of note, the methods differ widely in view of their computing time (Figure \ref{fig:figure8}B) and we saw no relationship between the accuracy of the estimates and the complexity of the method. Reasoning that over-estimating the number of dimensions is less problematic than under-estimating it, we kept the former methods for a full analysis of their impact on clustering (Figure \ref{fig:figure8}C-D), when combined with sctransform or the standard Seurat normalization. Although most methods performed well on the various clustering measures, the global maximum likelihood based on translated Poisson Mixture Model (\textit{maxLikGlobal}) provided the dimensionality estimate that best separated the subpopulations (Figure \ref{fig:figure8}C) and resulted in the best clustering accuracy (Figure \ref{fig:figure8}D). This method systematically selected many more components than were associated with the subpopulations, suggesting that although these additional components appear individually uninformative, in combination they nevertheless contribute to classification.

@@

For the methods that produce local dimensionality estimates, we used the maximum. For the elbow method, we implemented an automatic procedure by taking the farthest point from a line drawn between the variance explained by the first and last (i.e. 50th) components calculated. For the \texttt{JackStraw} method, since the \textit{Seurat} documentation advises not to use a p-value threshold (which would typically yield a very large number of dimensions) but rather look for a drop in significance, we applied the same farthest point algorithm on the log10(p-values), which in our hands reproduced manual threshold selection.

>>>>

Here, we develop a flexible R framework for pipeline comparison and evaluate the various steps of analysis leading from an initial count matrix to a cluster assignment, which are critical in a wide range of applications. We collected real datasets of known cell composition (Table \ref{tab:table1}) and used a variety of evaluation metrics to investigate in a multilevel fashion the impact of various parameters and variations around a core scRNAseq pipeline. Although we use some datasets based on other protocols, our focus is especially on droplet-based datasets that do not include exogenous control RNA (i.e., spike-ins); see Table \ref{tab:table1} and Figure \ref{fig:figure1} for more details. In addition to previously-used benchmark datasets with true cell labels \cite{duoClustering2018,tianMixology2018}, we simulated two datasets with a hierarchical subpopulation structure based on real 10x human and mouse data using \textit{muscat} \cite{CrowellMuscat2019}. 
Since graph-based clustering \cite{satijaSeurat2015} was previously shown to consistently perform well across several datasets \cite{duoClustering2018,tianMixology2018}, we used the \textit{Seurat} pipeline as the starting point to perform an integrated investigation of: 1) doublet identification, 2) cell filtering, 3) normalization, 4) feature selection, 5) dimension reduction, 6) clustering. We compared competing approaches and also explored more fine-grained parameters and variations on common methods. Importantly, the success of methods at a certain analytical step might be dependent on choices at other steps. Therefore, instead of evaluating each step in isolation, we developed a general framework for evaluating nested variations on a pipeline and suggest a multilevel panel of metrics. Finally, we evaluate several recent methods and provide concrete recommendations.

@@

Doublets, defined as two cells sequenced under the same cellular barcode (e.g., being captured in the same droplet), are fairly frequent in scRNAseq datasets, with estimates ranging from 1 to 10\% depending on the platform and cell concentration used \cite{bloomEstimating2018,kangMultiplexedDemuxlet2018}. While doublets of the same cell type are relatively innocuous for most downstream applications due to their conservation of the relative expression between genes, doublets formed from different cell types or states are likely to be misclassified and could potentially distort downstream analysis. In some cases, doublets can be identified through their unusually high number of reads and detected features, but this is not always the case (Supplementary Figure 2). A number of methods were developed to identify doublets, in most cases by comparing each cell to artificially-created doublets \cite{mcginnisDoubletfinder2019, LunScran2016, BaisScds2019}. We first evaluated the capacity of these methods to detect doublets using the two 10x datasets with cells of different genetic identity \cite{tianMixology2018}, using SNP genotypes as the ground truth. For the sole purpose of this section, we included an additional dataset with SNP information but lacking true cell labels \cite{kangMultiplexedDemuxlet2018}. Of note, SNP-based analyses also call doublets created by cells of the same cell type (but from different individuals) and which are generally described as homotypic (as opposed to neotypic or heterotypic doublets, i.e., doublets from different cell types). These homotypic doublets might not be identifiable from the mere gene counts, and their identification is not generally the primary aim of doublet callers since they are often considered innocuous and, when across individuals, can be identified through other means (e.g., SNPs). We therefore do not expect a perfect accuracy in datasets involving cells of the same type across individuals (as in the demuxlet dataset).

@@

We next investigated the impact of filtering according to various criteria (see Methods). An important risk of excluding cells on the basis of their distance from the whole distribution of cells on some properties (e.g., library size) is that these properties tend to have different distributions across subpopulations. As a result, thresholds in terms of number of MADs from the whole distribution can lead to strong biases against certain subpopulations (Figure \ref{fig:figure4}A). We therefore examined the tradeoff between the increased accuracy of filtering and the maximum proportion of cells excluded per subpopulation (Figure \ref{fig:figure4}B and Supplementary Figure 8). Since filtering changes the relative abundance of the different subpopulations, global clustering accuracy metrics such as ARI are not appropriate here. We therefore calculated the per-subpopulation precision and recall using the Hungarian algorithm \cite{PapadimitriouHu1998} and monitored their mean F1 score. A first observation was that, although the stringent filtering tended to be associated with an increase in accuracy, it could also become deleterious and most of the benefits could be achieved without very stringent filtering and minimizing subpopulation bias (Figure \ref{fig:figure4}B). Applying the same filtering criteria on individual clusters of cells (identified through scran's \texttt{quickCluster} method) resulted in nearly no cell being filtered out. This suggests that filtering on the global population tends to discard cells of subpopulations with more extreme properties (e.g., high library size), rather than low-quality cells. Finally, by changing the distributions, the doublet removal step in conjunction with filtering sometimes resulted in a net decrease in the proportion of excluded cells while retaining or improving accuracy. We therefore recommend the use of doublet removal followed by relatively mild filtering, such as is implemented in our `default' filtering (see Methods).

@@

We compared the various estimates of dimensionality in their ability to retrieve the intrinsic number of dimensions in a dataset, based on Seurat's weighted PCA space. As a first approximation of the true dimensionality, we computed the variance in each principal component that was explained by the subpopulations, which sharply decreased after the first few components in most datasets (Figure \ref{fig:figure8}A). Figure \ref{fig:figure8}B shows the difference between the dimension estimates of the above methods and that based on the subpopulations (i.e., from Figure \ref{fig:figure8}A). Of note, the methods differ widely in view of their computing time (Figure \ref{fig:figure8}B) and we saw no relationship between the accuracy of the estimates and the complexity of the method. Reasoning that over-estimating the number of dimensions is less problematic than under-estimating it, we kept the former methods for a full analysis of their impact on clustering (Figure \ref{fig:figure8}C-D), when combined with sctransform or the standard Seurat normalization. Although most methods performed well on the various clustering measures, the global maximum likelihood based on translated Poisson Mixture Model (\textit{maxLikGlobal}) provided the dimensionality estimate that best separated the subpopulations (Figure \ref{fig:figure8}C) and resulted in the best clustering accuracy (Figure \ref{fig:figure8}D). This method systematically selected many more components than were associated with the subpopulations, suggesting that although these additional components appear individually uninformative, in combination they nevertheless contribute to classification.

@@

For the methods that produce local dimensionality estimates, we used the maximum. For the elbow method, we implemented an automatic procedure by taking the farthest point from a line drawn between the variance explained by the first and last (i.e., 50th) components calculated. For the \texttt{JackStraw} method, since the \textit{Seurat} documentation advises not to use a p-value threshold (which would typically yield a very large number of dimensions) but rather look for a drop in significance, we applied the same farthest point algorithm on the log10(p-values), which in our hands reproduced manual threshold selection.
